Ecosystem size and complexity are extrinsic drivers of food chain length in branching networks

### 2022-03-30 rmd/manuscript.Rmd
edit: manuscript

In ecosystems with a low disturbance regime, FCL was largely determined by omnivory with limited effects of ecosystem size and complexity (colored lines in **Figure 2**). Unsurprisingly, FCL was negatively related to the degree of omnivory, with no-omnivory food chains ($P$ only consumes $C$) having the longest FCL, and strong omnivory ($P$ consumes very little $C$) having the shortest FCL. These trends were consistent across dispersal scenarios (**Figure 2**).

@@

A high disturbance regime produced much different responses of FCL. FCL increased across both gradients of ecosystem size and complexity under all combinations of productivity, the degree of omnivory, and dispersal distance (**Figure 3**). However, the effects of ecosystem size and complexity were mediated by ecological factors differently. Ecosystem size effects were stronger with longer dispersal distance ($\theta = 0.1$), while the opposite was true for ecosystem complexity (**Figure 3**). In addition, the effect of ecosystem complexity was pronounced in food webs with weaker omnivory (**Figure 3B**).

Interestingly, the positive effects of ecosystem size and complexity were driven by different mechanisms. For ecosystem size, the underlying mechanism was consistent when productivity was high; increased occupancies of all the three species resulted in longer food chains. However, when productivity was low, the mechanisms were more complex, particularly with strong omnivory (the third column in **Figure S1 and S2**). In this scenario, the occupancies of $B$ and $C$ decreased at larger ecosystem sizes, while $P$ increased its occupancy (**Figure S1 and S2**). Therefore, $P$ compensated for the loss of $B$ and $C$ to maintain the positive relationship between FCL and ecosystem size. For ecosystem complexity, all the three species increased with ecosystem complexity when productivity was high, translating into longer food chains in more complex ecosystems (**Figure S3 and S4**). However, when food webs contained strong omnivory with low productivity (the third column in **Figure S3 and S4**), the increased occupacy of $C$ maintained the positive relationship between FCL and ecosystem complexity by offsetting the loss of $P$. Hence, the contrasting mechanisms underlie the positive effects of ecosystem size and complexity on FCL.

@@

Discussion to be added

>>>>

In ecosystems with a low disturbance regime, FCL was largely determined by omnivory with limited effects of ecosystem size and complexity (colored lines in **Figure 2**). Unsurprisingly, FCL was negatively related to the degree of omnivory, with no-omnivory food chains ($P$ only consumes $C$) having the longest FCL, and strong omnivory ($P$ consumes very little $C$) having the shortest FCL. In addition, productivity played a secondary role in increasing FCL, particularly when omnivory was strong (compare left and right panels in **Figure 2**). Dispersal had no apparent effects on FCL (compare top and bottom panels in **Figure 2**). These trends were consistent across dispersal scenarios (**Figure 2**).

@@

On average, increased disturbance frequency caused the dominance of short food chains (**Figures 2 and 3**). More importantly, however, a high disturbance regime produced much different responses of FCL to ecosystem properties. FCL increased across both gradients of ecosystem size and complexity under all combinations of productivity, the degree of omnivory, and dispersal distance (**Figure 3**). The effect of ecosystem complexity was generally clearer than that of ecosystem size. It is important to note that the positive relationships emerged in a high disturbance regime because smaller and simpler ecosystems were more vulnerable to increased disturbance frequency (compare far left in **Figures 2 and 3**).

Dispersal buffered the reduction of FCL due to increased disturbance, but it influenced FCL differently across the gradients of ecosystem structure. The greater buffering effect of dispersal emerged in larger ecosystems (compare top and bottom panels in **Figure 3**) because greater dispersal capability is required to reach distant patches. In the meantime, simpler ecosystems benefitted more from increased dispersal because habitat patches are more isolated from one another in linearly arranged networks (**Figure 3**). Productivity increased FCL as observed in a low disturbance regime, but it did not modulate the qualitative relationships between FCL and ecosystem properties (compare left and right panels in **Figure 3**).

Interestingly, the positive effects of ecosystem size and complexity were driven by different mechanisms. For ecosystem size, the underlying mechanism was consistent when productivity was high; increased occupancies of all the three species resulted in longer food chains (i.e., "additive mechanism"). However, when productivity was low, the mechanisms were more complex, particularly with strong omnivory (the third column in **Figure S1 and S2**). In this scenario, the occupancies of $B$ and $C$ decreased at larger ecosystem sizes, while $P$ increased its occupancy ("insertion mechanism"; **Figure S1 and S2**). Therefore, $P$ compensated for the loss of $B$ and $C$ to maintain the positive relationship between FCL and ecosystem size. For ecosystem complexity, all the three species increased with ecosystem complexity when productivity was high, translating into longer food chains in more complex ecosystems (**Figure S3 and S4**). However, when food webs contained strong omnivory with low productivity (the third column in **Figure S3 and S4**), the increased occupacy of $C$ maintained the positive relationship between FCL and ecosystem complexity by offsetting the loss of $P$. Hence, the contrasting structural mechanisms underlie the positive association of ecosystem size and complexity on FCL.

@@

The ecosystem size hypothesis of food chain length has received support from both theoretical and empirical research [@vanderzandenPatternsFoodChain1999; @postEcosystemSizeDetermines2000; @takimotoEffectsProductivityDisturbance2012], although empirical work so far has rarely elaborated the underlying drivers [but see @saboRoleDischargeVariation2010; @wardMechanisticTheoryAquatic2017]. In contrast, while the importance of ecosystem complexity is increasingly recognized in ecology [@grantLivingBranchesPopulation2007; @altermattDiversityRiverineMetacommunities2013; @teruiMetapopulationStabilityBranching2018; @teruiEmergentDualScaling2021], it has not been incorporated in theoretical explorations of food chain length to date. Here, we show that food chain length responds positively to both increasing ecosystem size and complexity. This supports our hypothesis that ecosystem complexity may represent an overlooked yet important extrinsic control of food chain length. However, the relationship between FCL and ecosystem complexity varied based on additional extrinsic and intrinsic factors, supporting the emerging paradigm that there is no one single driver of FCL in ecosystems [@mchughDualInfluencesEcosystem2010; @wardMechanisticTheoryAquatic2017]. Although we focus on river networks here, we expect our results to be transferable to other branching structures, such as mountain ranges and cave systems.

While increased disturbance frequency consistently reduced food chain length as predicted by dynamical stability hypothesis [@pimmNumberTrophicLevels1977], more important findings in this study were evident from their synergistic effects with ecosystem structure. Our low disturbance regime scenario can be viewed as habitat networks having nearly homogeneous environments, in which long food chains were generally supported independent of ecosystem size and complexity. Consequently, FCL was predominantly determined by food web structure. However, under high disturbance regimes, only large and complex networks could provide diverse and heterogeneous habitats with potential refugia, allowing species to thrive and maintain long food chains. While empirical studies rarely looked at the influence of ecosystem structure on FCL, several studies provide some support for this theoretical prediction. For example, @sullivanEcosystemStructureEmerges2015 reported FCL is controlled parimarily by the number of tributary confluences in Idaho streams. Similarly, it has been shown that the merging of major tributaries lengthened FCL in the Colorado River, especially during a post-monsoon turbid water phase [@saboPulsedFlowsTributary2018]. Clearly, more empirical research is needed to confirm our theoretical predictions.

Although dispersal has received little attention in food chain research, our simulation highlighted its importance in regulating food chain length. In particular, high dispersal was required for large, linearly-arranged ecosystems to maintain long food chains under high disturbance regimes. This is probably because sufficient propagules are needed to successfully recolonize after disturbance. Human activities that affect habitat connectivity and dispersal ability of species within riverine networks may compromise the buffering effects of ecosystem size and complexity against disturbance by limiting dispersing propagules. For example, dams and diversion have altered the connectivity between suitable habitat patches, or have completely blocked dispersal ability 46,47. Light pollution has also negatively affected the ability of adult aquatic insects to disperse between habitat patches 48.

The interactive effect of disturbance and ecosystem structure has important implications for ecosystem management because anthropogenic climate change is predicted to increase the frequency and magnitude of disturbances [@brunnerFutureStreamflowRegime2020; @mittalImpactHumanIntervention2016]. River networks are especially vulnerable due to their propensity to go dry during droughts [@zipperPervasiveChangesStream2021] as well as susceptibility to more intense flood disturbance with increasing storm intensity [@schiermeierIncreasedFloodRisk2011]. However, human activities have compromised the habitat heterogeneity and structural connectivity within riverine networks [@ishiyamaPredictingEcologicalImpacts2018; @moritaEffectsHabitatFragmentation2002; @barbarossaImpactsCurrentFuture2020]. Our theory implies that these human-induced alterations that simplfy river networks may make riverine food webs more vulnerable to the anticipated environmental changes.

The context dependency of disturbance, productivity and dispersal has important implications for the conservation and management of freshwater ecosystems. This adds an important distinction in the discussion of the suitability of assigning protection or conservation status to regions. Current conservation schemes generally emphasize size, aiming to protect the largest areas possible. While this is certainly a useful attribute generally, designating regions with a high degree of variability, or many small patches, may be more beneficial than a large, homogenous area 49, and accounting for ecosystem complexity is particularly suited to ecosystem architectures in which branches have variable environmental conditions. Conservation and management activities should focus on activities that increase the habitat heterogeneity and connectivity of suitable habitat patches to increase resilience to predicted increases in disturbance frequency and magnitude.

The role of food web motifs on FCL largely followed our predictions (**Figure 1**). Specifically, FCL declined as the value of the resource B, and thus the proportion in the diet, to the predator P increased. Essentially, the relative qualitative values of the resources to the predator controlled the level of omnivory and FCL. Although FCL was greatest when the community was arranged as a simple linear food chain, this arrangement was not the most stable. Specifically, when disturbance regime was high, and productivity and dispersal ability were low, the observed FCL across networks was \~1 (although increasing dispersal ability and/or carrying capacity resulted in food chains of \~2.5-3). Therefore, in the absence of prey resource choice, strong, top-down trophic interactions were destabilizing, leading to the extirpation of consumer species at higher trophic levels across the networks.

Changes in community composition are occurring worldwide and in all habitats. Species extirpations and extinctions, coupled with species introductions and invasions are creating novel communities worldwide. In Lake Tahoe (USA), for example, replacement of the top fish predator and the introduction of a freshwater shrimp substantially altered the food web structure 50. Understanding the interactions within these novel communities, and how this could affect food chain lengths is an important goal of contemporary ecologists. The strength of omnivory in these novel communities can be increased or decreased, depending on the introduced species and the extant species pool or community. For example, non-indigenous fishes in the Eastern Mediterranean Sea outcompete native species for high quality prey items, causing native species to increase consumption of non-preferred prey items51. Introductions of non-native fishes in the Colorado River basin have led to longer FCL and narrower niches (e.g., lower degree of omnivory) of native species52. Alternatively, the introduction of hyper-successful basal or primary consumer species can weaken or remove the omnivory link by becoming the dominant prey resource. For example, nonnative New Zealand mud snail dominated the macroinvertebrate biomass in streams in Yellowstone National Park and likely affected food web structure53.  Having a single prey resource dominate the community is akin to reducing the degree of omnivory (i.e., removing prey choice), which destabilizes the community, particularly under high disturbance regimes.

It is abundantly clear that ecosystem complexity in natural systems is the rule, not the exception. Early theoretical and empirical work has assumed simple habitat structure as a necessity. However, habitat heterogeneity, such as the presence of prey refugia, and dispersal between connected habitat patches allow communities to stabilize and persist through space and time 54--56. Complex ecosystem structure is an extrinsic factor that can affect food chain length in communities. Incorporating a realistic understanding of ecosystem structural complexity could alter our perception of communities in naturally complex systems, potentially leading to paradigm shifts in macroecology.
